---
title: "benviplot Visual Testing"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  warning: false
  message: false
  fig-width: 8
  fig-height: 5
  fig-align: center
---

# Setup

```{r setup}
library(benviplot)
library(ggplot2)
library(dplyr)

# Load package data
data(iqa)
data(iqaiw)
```

# Margin Testing

Test the updated margins in `theme_benvi()` - title, subtitle, and caption spacing.

```{r margin-test}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = benvi_palette("benvi_blue")[3]) +
  labs(
    title = "Car Weight vs Fuel Efficiency",
    subtitle = "Testing margin spacing between title, subtitle, and caption",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    caption = "Source: Motor Trend 1974 dataset"
  ) +
  theme_benvi()
```

# Line Plots

## Single Line (economics)

```{r line-single}
plot_line(economics, x = date, y = unemploy / 1000) +
  labs(
    title = "US Unemployment Over Time",
    subtitle = "Single line plot with standard dataset",
    y = "Unemployed (millions)",
    x = "Date"
  )
```

## Multiple Lines (iqaiw)

```{r line-multiple}
# Filter for 3 major cities
iqaiw_subset <- iqaiw |>
  filter(
    name_muni %in% c("São Paulo", "Rio de Janeiro", "Belo Horizonte"),
    rooms == "Total"
    )

plot_line(
  iqaiw_subset,
  x = date,
  y = index,
  variable = name_muni,
  palette = "qual_5",
  zero = FALSE
) +
  labs(
    title = "IQAIW: Rental Price Index",
    subtitle = "Using internal iqaiw dataset with city grouping",
    x = "Date",
    y = "Index (base = 100)"
  )
```

## Line with Points

```{r line-points}
plot_line(
  economics |> filter(date >= as.Date("2010-01-01")),
  x = date,
  y = psavert,
  point = TRUE
) +
  labs(
    title = "Personal Savings Rate (2010-2015)",
    subtitle = "Line plot with point markers",
    y = "Personal Savings Rate (%)",
    x = "Date"
  )
```

# Column Plots

## Basic Column (iqa)

```{r column-basic}
# Recent data from major cities
date_recent <- max(iqa$date, na.rm = TRUE)

iqa_recent <- iqa |>
  filter(date == max(date)) |>
  slice_max(price_m2, n = 5)

plot_column(iqa_recent, x = name_muni, y = price_m2) +
  labs(
    title = "Median Rent by City",
    subtitle = paste0(
      "Using internal IQA dataset - ",
      format(date_recent, format = "%b/%Y")
    ),
    x = NULL,
    y = "Rent per m² (R$)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Column with Text Labels

```{r column-text}
# Sample data
products <- data.frame(
  product = c("A", "B", "C", "D", "E"),
  sales = c(120, 200, 150, 180, 220)
)

plot_column(
  products,
  x = product,
  y = sales,
  fill = benvi_palette("purples")[1],
  text = TRUE
  ) +
  labs(
    title = "Product Sales with Value Labels",
    subtitle = "Testing text parameter",
    x = "Product",
    y = "Sales (units)"
  )
```

## Grouped Columns

```{r column-grouped}
# Quarterly data with grouping
quarterly <- data.frame(
  quarter = rep(c("Q1", "Q2", "Q3", "Q4"), each = 3),
  region = rep(c("North", "South", "West"), 4),
  revenue = c(120, 130, 115, 145, 155, 140, 160, 165, 150, 175, 180, 170)
)

plot_column(
  quarterly,
  x = quarter,
  y = revenue,
  variable = region,
  palette = "greens",
  scale_name = "Region"
) +
  labs(
    title = "Quarterly Revenue by Region",
    subtitle = "Grouped column chart",
    x = "Quarter",
    y = "Revenue ($1000s)"
  )
```

# Scatter Plots

## Basic Scatter (mtcars)

```{r scatter-basic}
plot_scatter(mtcars, x = wt, y = mpg) +
  labs(
    title = "Car Weight vs MPG",
    subtitle = "Simple scatter plot",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )
```

## Grouped Scatter (iris)

```{r scatter-grouped}
plot_scatter(
  iris,
  x = Sepal.Length,
  y = Sepal.Width,
  variable = Species,
  palette = "qual_7",
  scale_name = "Species"
) +
  labs(
    title = "Iris Sepal Dimensions by Species",
    subtitle = "Using iris dataset with species grouping",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  )
```

## Scatter with Fit Line (mtcars)

```{r scatter-fit}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  fit = TRUE,
  fit_method = "lm"
) +
  labs(
    title = "Weight vs MPG with Linear Trend",
    subtitle = "Testing fit parameter",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )
```

## Grouped Scatter with Grouped Fits

```{r scatter-fit-grouped}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  variable = as.factor(cyl),
  fit = TRUE,
  fit_variable = TRUE,
  fit_method = "lm",
  palette = "purples",
  scale_name = "Cylinders"
) +
  labs(
    title = "Weight vs MPG by Cylinder Count",
    subtitle = "Grouped scatter with separate trend lines",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )
```

# Area Plots

## Single Area (economics)

```{r area-single}
plot_area(economics, x = date, y = unemploy / 1000) +
  labs(
    title = "US Unemployment Over Time",
    subtitle = "Simple area chart",
    x = "Date",
    y = "Unemployed (millions)"
  )
```

## Stacked Area (iqaiw)

```{r area-stacked}
# Use 3 cities for stacked area
iqaiw_cities <- iqaiw |>
  filter(name_muni %in% c("São Paulo", "Rio de Janeiro", "Belo Horizonte"))

plot_area(
  iqaiw_cities,
  x = date,
  y = price_m2,
  variable = name_muni,
  palette = "seq_blues",
  scale_name = "City"
) +
  labs(
    title = "Rental Price per m² by City",
    subtitle = "Stacked area using iqaiw dataset",
    x = "Date",
    y = "Price (R$/m²)"
  )
```

# Histograms

## Basic Histogram (mtcars)

```{r histogram-basic}
plot_histogram(mtcars, x = mpg) +
  labs(
    title = "Distribution of MPG",
    subtitle = "Simple histogram",
    x = "Miles per Gallon",
    y = "Count"
  )
```

## Histogram with Custom Bins

```{r histogram-bins}
plot_histogram(mtcars, x = mpg, bins = 15) +
  labs(
    title = "Distribution of MPG (15 bins)",
    subtitle = "Testing bins parameter",
    x = "Miles per Gallon",
    y = "Count"
  )
```

## Grouped Histogram (iris)

```{r histogram-grouped}
plot_histogram(
  iris,
  x = Sepal.Length,
  variable = Species,
  palette = "oranges",
  scale_name = "Species",
  bins = 20
) +
  labs(
    title = "Distribution of Sepal Length by Species",
    subtitle = "Grouped histogram using iris dataset",
    x = "Sepal Length (cm)",
    y = "Count"
  )
```

# Palette Testing

## Theme Palettes (4 colors)

```{r palette-theme}
sample_data <- data.frame(
  category = factor(c("A", "B", "C", "D")),
  value = c(25, 40, 35, 30)
)

ggplot(sample_data, aes(x = category, y = value, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "purples") +
  labs(
    title = "Theme Palette: Purples",
    subtitle = "4-color theme palette",
    x = NULL,
    y = "Value"
  ) +
  theme_benvi()
```

## Qualitative Palettes (8 colors)

```{r palette-qual}
# Create data with more categories
qual_data <- data.frame(
  category = factor(LETTERS[1:6]),
  value = c(23, 45, 32, 18, 38, 42)
)

ggplot(qual_data, aes(x = category, y = value, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "qual_5") +
  labs(
    title = "Qualitative Palette: qual_5",
    subtitle = "8-color qualitative palette",
    x = NULL,
    y = "Value"
  ) +
  theme_benvi()
```

## Sequential Palette (continuous)

```{r palette-sequential}
# Heatmap data
set.seed(123)
heatmap_data <- expand.grid(
  x = 1:10,
  y = 1:10
)
heatmap_data$value <- with(heatmap_data, x * y + rnorm(100, 0, 5))

ggplot(heatmap_data, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_benvi_c(pal_name = "seq_greens", name = "Value") +
  labs(
    title = "Sequential Palette: seq_greens",
    subtitle = "9-color sequential palette for continuous data"
  ) +
  theme_benvi() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

## City Palettes

```{r palette-city}
# Rio palette
city_data <- data.frame(
  district = c("Centro", "Zona Sul", "Zona Norte", "Zona Oeste"),
  price = c(5500, 7200, 3800, 4100)
)

ggplot(city_data, aes(x = district, y = price, fill = district)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "rio_qual") +
  labs(
    title = "City Palette: Rio de Janeiro",
    subtitle = "rio_qual palette for city-specific visualizations",
    x = NULL,
    y = "Rent (R$)"
  ) +
  theme_benvi() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## São Paulo Palette

```{r palette-spo}
ggplot(city_data, aes(x = district, y = price, fill = district)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "spo_qual") +
  labs(
    title = "City Palette: São Paulo",
    subtitle = "spo_qual palette",
    x = NULL,
    y = "Rent (R$)"
  ) +
  theme_benvi() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Multiple Plots Grid

Comparing different themes side by side:

```{r grid-comparison, fig.width=10, fig.height=8}
library(patchwork)

p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("blues")[3], size = 2) +
  labs(title = "Scatter Plot", subtitle = "theme_benvi()") +
  theme_benvi()

p2 <- plot_line(economics |> filter(date >= as.Date("2010-01-01")),
                x = date, y = unemploy / 1000) +
  labs(title = "Line Plot", subtitle = "Standard benvi theme")

p3 <- plot_column(products, x = product, y = sales, text = TRUE) +
  labs(title = "Column Plot", subtitle = "With text labels")

p4 <- plot_histogram(iris, x = Sepal.Length, variable = Species,
                     palette = "greens", scale_name = "Species") +
  labs(title = "Histogram", subtitle = "Grouped by species")

(p1 | p2) / (p3 | p4)
```

# Summary

This document provides a comprehensive visual test suite for the benviplot package, including:

- **Margin testing**: Verifying title/subtitle/caption spacing
- **Plot functions**: All helper functions (line, column, scatter, area, histogram)
- **Datasets**: Both internal (iqa, iqaiw) and standard R datasets (mtcars, iris, economics)
- **Palettes**: Theme, qualitative, sequential, and city-specific palettes
- **Theme consistency**: Testing theme_benvi() across various plot types

Use this document to quickly evaluate visual changes after package updates.
