---
title: "Color Palettes in benviplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Color Palettes in benviplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ggplot2)
library(benviplot)
```

## Introduction

This vignette provides a comprehensive guide to all color palettes available in `benviplot`. You'll learn about the different palette types, when to use each one, and see visual examples of every palette.

## Palette Types

`benviplot` organizes palettes into several categories based on their intended use:

- **Basic Palettes** (3 colors): Simple, versatile palettes
- **Set Palettes** (4 colors): Small categorical datasets
- **Qualitative Palettes** (8 colors): Larger categorical datasets
- **Sequential Palettes** (9 colors): Continuous/ordered data
- **City-Specific Palettes**: Themed palettes for Brazilian cities
- **Index Palettes**: Specialized single-color scales

## Basic Palettes

Three fundamental palettes with 3 colors each:

```{r basic-palettes}
# Basic palette (3 colors)
benvi_palette("basic")
```

**When to use**: Simple visualizations with 2-3 categories.

**Example**:

```{r basic-example}
data <- data.frame(
  category = c("Low", "Medium", "High"),
  value = c(23, 45, 32)
)

ggplot(data, aes(x = category, y = value, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "basic") +
  labs(title = "Using Basic Palette", x = NULL, y = "Value") +
  theme_benvi()
```

## Set Palettes (Set0 - Set7)

Eight themed palettes, each with 4 carefully selected colors. Ideal for small categorical datasets.

```{r set-palettes-display}
# Display all theme palettes
theme_names <- c("grays", "browns", "yellows", "greens", "blues", "purples", "pinks", "oranges")

for (pal_name in theme_names) {
  print(benvi_palette(pal_name))
}
```

**When to use**:
- Categorical data with 2-4 groups
- Charts where you need strong color contrast
- Professional presentations

**Example**:

```{r set-example}
# Sample data with 4 regions
regions <- data.frame(
  region = c("North", "South", "East", "West"),
  sales = c(120, 150, 130, 160)
)

ggplot(regions, aes(x = region, y = sales, fill = region)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "purples") +
  labs(title = "Regional Sales", x = NULL, y = "Sales ($1000s)") +
  theme_benvi()
```

## Qualitative Palettes (Qual1 - Qual9)

Nine palettes with 8 distinct colors each. Designed for larger categorical datasets where you need many distinguishable colors.

```{r qual-palettes-display}
# Display all Qualitative palettes
qual_names <- paste0("qual_", 1:9)

for (pal_name in qual_names) {
  print(benvi_palette(pal_name))
}
```

**When to use**:
- Categorical data with 5-8 groups
- Complex visualizations with many categories
- Dashboards and multi-series charts

**Example**:

```{r qual-example}
# Use mtcars with multiple groups
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(carb))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "qual_3", name = "Carburetors") +
  labs(
    title = "MPG vs Weight by Carburetor Count",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

## Sequential Palettes (Seq0 - Seq7)

Eight gradient palettes with 9 colors each, transitioning from light to dark. Perfect for continuous or ordered data.

```{r seq-palettes-display}
# Display all Sequential palettes
seq_names <- c("seq_grays", "seq_browns", "seq_yellows", "seq_greens", "seq_blues", "seq_purples", "seq_pinks", "seq_oranges")

for (pal_name in seq_names) {
  print(benvi_palette(pal_name))
}
```

**When to use**:
- Continuous numerical data
- Heatmaps and choropleth maps
- Ordered categorical data (e.g., Low/Medium/High)
- Data where progression/intensity matters

**Example**:

```{r seq-example}
# Create a heatmap
set.seed(123)
heatmap_data <- expand.grid(
  x = 1:10,
  y = 1:10
)
heatmap_data$value <- with(heatmap_data, x * y + rnorm(100, 0, 5))

ggplot(heatmap_data, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_benvi_c(pal_name = "seq_greens", name = "Value") +
  labs(title = "Heatmap with Sequential Palette") +
  theme_benvi() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

## City-Specific Palettes

Themed palettes for three major Brazilian cities. Each city has qualitative, sequential, and diverging variants.

### São Paulo Palettes

```{r spo-palettes}
benvi_palette("spo_seq")
benvi_palette("spo_div")
benvi_palette("spo_qual")
```

### Rio de Janeiro Palettes

```{r rio-palettes}
benvi_palette("rio_seq")
benvi_palette("rio_div")
benvi_palette("rio_qual")
```

### Belo Horizonte Palettes

```{r bhe-palettes}
benvi_palette("bhe_seq")
benvi_palette("bhe_div")
```

**When to use**:
- Geographic visualizations (maps, regional data)
- Reports specific to these cities
- When you want location-themed aesthetics

**Example**:

```{r city-example}
# Sample city data
city_data <- data.frame(
  district = c("Centro", "Zona Sul", "Zona Norte", "Zona Oeste"),
  price = c(5500, 7200, 3800, 4100)
)

ggplot(city_data, aes(x = district, y = price, fill = district)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "rio_qual") +
  labs(
    title = "Average Rent by District - Rio de Janeiro",
    x = NULL,
    y = "Price (R$)"
  ) +
  theme_benvi() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Index Palettes

Specialized single-color gradients for index visualizations.

```{r index-palettes}
# Blue index
benvi_palette("benvi_blue")

# Purple index
benvi_palette("benvi_purple")
```

**When to use**:
- Financial indices and time series
- Single-metric dashboards
- When you want monochromatic emphasis

**Example**:

```{r index-example}
# Stock index over time
index_data <- data.frame(
  date = seq.Date(as.Date("2020-01-01"), as.Date("2024-12-31"), by = "month"),
  index = 100 + cumsum(rnorm(60, 0.5, 3))
)

ggplot(index_data, aes(x = date, y = index)) +
  geom_line(color = benvi_palette("benvi_blue")[7], linewidth = 1) +
  geom_area(fill = benvi_palette("benvi_blue")[3], alpha = 0.3) +
  labs(
    title = "Market Index Over Time",
    x = "Date",
    y = "Index Value"
  ) +
  theme_benvi()
```

## Advanced Usage

### Reversing Palettes

Any palette can be reversed using `direction = -1`:

```{r reverse}
# Original
benvi_palette("seq_purples")

# Reversed
benvi_palette("seq_purples", direction = -1)
```

This is useful when you want dark-to-light instead of light-to-dark:

```{r reverse-example}
ggplot(heatmap_data, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_benvi_c(pal_name = "seq_purples", direction = -1, name = "Value") +
  labs(title = "Reversed Sequential Palette") +
  theme_benvi() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

### Extracting Specific Colors

Get individual colors from a palette:

```{r extract-colors}
# Get the 3rd color from Qual5
my_color <- benvi_palette("qual_5")[3]
my_color

# Use in a plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = my_color, size = 3) +
  labs(title = "Using a Single Extracted Color") +
  theme_benvi()
```

### Custom Number of Colors

For discrete palettes, request fewer colors:

```{r custom-n}
# Get only 4 colors from an 8-color palette
benvi_palette("qual_7", n = 4)
```

For continuous palettes, interpolate any number:

```{r interpolate}
# Interpolate to get 15 colors
benvi_palette("seq_yellows", n = 15, type = "continuous")
```

### Combining Palettes

Mix colors from different palettes for custom schemes:

```{r combine}
# Create custom palette from multiple sources
custom_colors <- c(
  benvi_palette("greens")[1],
  benvi_palette("qual_5")[4],
  benvi_palette("rio_qual")[2],
  benvi_palette("seq_oranges")[8]
)

# Use in plot
sample_data <- data.frame(
  category = LETTERS[1:4],
  value = c(25, 40, 35, 30)
)

ggplot(sample_data, aes(x = category, y = value, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = custom_colors) +
  labs(title = "Custom Combined Palette") +
  theme_benvi()
```

## Accessibility Considerations

When choosing palettes, consider:

1. **Color blindness**: Test palettes with color blindness simulators
2. **Print compatibility**: Sequential palettes work best in grayscale
3. **Contrast**: Ensure sufficient contrast for readability
4. **Number of colors**: Don't use more categories than necessary

### Tips for Accessibility

- **Limit categories**: Use 5-7 colors maximum when possible
- **Add patterns**: Combine colors with line types or shapes
- **Use labels**: Don't rely solely on color to convey information

```{r accessibility}
# Example with both color and shape
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl), shape = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "purples", name = "Cylinders") +
  scale_shape_manual(values = c(16, 17, 15), name = "Cylinders") +
  labs(
    title = "Using Color AND Shape for Accessibility",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

## Palette Selection Guide

Quick reference for choosing the right palette:

| Data Type | Palette Type | Examples | Color Count |
|-----------|-------------|----------|-------------|
| 2-3 categories | Basic | basic | 3 |
| 2-4 categories | Theme | grays, browns, yellows, etc. | 4 |
| 5-8 categories | Qualitative | qual_1, qual_2, ..., qual_9 | 8 |
| Continuous data | Sequential | seq_grays, seq_blues, etc. | 9 (interpolated) |
| Geographic (cities) | City-specific | spo_*, rio_*, bhe_* | 4-8 |
| Single metric | Brand | benvi_blue, benvi_purple | 9 |

## Complete Palette Gallery

Here's a quick visual reference of all available palettes:

```{r gallery, fig.height=12}
# Create a function to display palette swatches
show_palette_grid <- function(palette_names, title) {
  n_pals <- length(palette_names)
  par(mfrow = c(ceiling(n_pals / 2), 2), mar = c(1, 4, 2, 1))

  for (pal_name in palette_names) {
    pal <- benvi_palette(pal_name)
    n_colors <- length(pal)

    plot(
      1:n_colors,
      rep(1, n_colors),
      col = pal,
      pch = 15,
      cex = 4,
      axes = FALSE,
      xlab = "",
      ylab = "",
      main = pal_name,
      xlim = c(0.5, n_colors + 0.5)
    )
  }

  mtext(title, outer = TRUE, line = -2, cex = 1.5, font = 2)
}

# Display theme palettes
show_palette_grid(c("grays", "browns", "yellows", "greens", "blues", "purples", "pinks", "oranges"), "Theme Palettes")
```

```{r gallery-qual, fig.height=12}
# Display Qualitative palettes
show_palette_grid(paste0("qual_", 1:9), "Qualitative Palettes")
```

```{r gallery-seq, fig.height=12}
# Display Sequential palettes
show_palette_grid(c("seq_grays", "seq_browns", "seq_yellows", "seq_greens", "seq_blues", "seq_purples", "seq_pinks", "seq_oranges"), "Sequential Palettes")
```

## Summary

In this vignette you learned:

- ✓ All available palette types and when to use them
- ✓ How to view and extract colors from palettes
- ✓ Advanced techniques (reversing, combining, interpolating)
- ✓ Accessibility considerations for color choices
- ✓ Quick reference guide for palette selection

## Next Steps

- See `vignette("plot-functions")` for using palettes with helper functions
- See `vignette("themes-and-styling")` for advanced theme customization
- Explore the function reference at `?benvi_palette` and `?scale_color_benvi_d`

## Further Reading

For more on color theory and data visualization:

- [Colorbrewer](https://colorbrewer2.org/) - Color advice for cartography
- [Adobe Color](https://color.adobe.com/) - Color wheel and palette tools
