---
title: "Font Setup Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Font Setup Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Starting with version 1.1.0, `benviplot` uses a modern font rendering system based on `systemfonts` and `ragg`. This change **eliminates the frustrating DPI issues** that plagued the previous showtext-based system, providing consistent, high-quality text rendering across all outputs.

## Why We Changed

The previous system (<=v1.0.0) used the `showtext` package, which had a DPI mismatch problem. As a result, users had to adjust `showtext_opts(dpi = ...)` settings before exporting plots, and output varied unpredictably across different devices and IDEs. Changing the DPI settings would also affect the resolution of the plots inside the RStudio Plots pane, making for a less than ideal user experience.

The new system fixes these issues. The only trade-off is that Poppins must be installed locally. Installing Poppins, however, is simple and free. Additionally, if Poppins is not available, the system will fall back to system "sans" font and will still render correctly at any resolution.

# Installation

The optimal setup requires: 1) installing Poppins; and  2) installing the `ragg` package. For RStudio users, an additional step is required to set AGG as the default graphics device.

### Install Poppins Font

The most reliable way to install Poppins is to get it from Google Fonts.

1. Go to [Google Fonts - Poppins](https://fonts.google.com/specimen/Poppins)
2. Click "Download family"
3. Extract the ZIP file
4. Install the fonts

Remember to restart R to make the fonts available.

An alternative way to install Poppins is to run `install_poppins()`. This function might not work on all systems, however.

```{r, eval=FALSE}
# Install Poppins font from Google Fonts
benviplot::install_poppins()
```

### Install ragg

For the best graphics quality, install the `ragg` package.

```{r, eval=FALSE}
install.packages("ragg")
```

Remember to restart R and/or your IDE to make sure everything works as intended.

---

### Configure RStudio (Optional)

To use ragg as the default graphics device in RStudio:

1. Go to **Tools → Global Options → General → Graphics**
2. Set **Backend** to **AGG**
3. Click **OK** and restart RStudio

This ensures all plots in the RStudio Plots pane use ragg for rendering.

---

### Quick Start: One-Command Setup

An alternative, but still experimental way to set up fonts for benviplot is to run `setup_benvi_fonts()`.

```{r, eval=FALSE}
library(benviplot)

# One-time setup - installs Poppins and provides RStudio configuration guidance
setup_benvi_fonts()
```

This function will:

1. Download and install the Poppins font to your system.
2. Check if `ragg` is installed (recommended for best quality).
3. Provide instructions for configuring RStudio.

## Check Your Setup

Use `font_status()` to check your current configuration:

```{r, eval=FALSE}
font_status()
```

# Using benviplot with Poppins

## Theme Behavior

The `theme_benvi()` function automatically detects whether Poppins is installed. If it's not installed, it falls back to system "sans" font.

```{r, eval=FALSE}
library(ggplot2)
library(benviplot)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "My Plot") +
  theme_benvi()  # Automatically uses Poppins or fallback
```

## Using Other Fonts

While benviplot is designed for Poppins, you can use any system font with the systemfonts framework.

```{r, eval=FALSE}
# List all available system fonts
systemfonts::system_fonts()

# Use a different font in theme_benvi
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_benvi() +
  theme(text = element_text(family = "Roboto"))  # Override font
```

# Summary

The new font system in benviplot v1.1.0 aims to provide a better user experience.

- **No DPI issues**: Consistent rendering at any resolution.
- **Better quality**: Superior text rendering with ragg.
- **Simpler code**: No per-session font loading required.
