---
title: "Font Setup Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Font Setup Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Starting with version 1.1.0, `benviplot` uses a modern font rendering system based on `systemfonts` and `ragg`. This change **eliminates the frustrating DPI issues** that plagued the previous showtext-based system, providing consistent, high-quality text rendering across all outputs.

## Why We Changed

The previous system (v1.0.0) used the `showtext` package, which had a notorious DPI mismatch problem:

- Text would render at different sizes in RStudio vs saved plots
- Users had to constantly adjust `showtext_opts(dpi = ...)` settings
- Output varied unpredictably across different devices
- No reliable one-size-fits-all solution

**The new system fixes all of this.** By using systemfonts and ragg, you get:

✅ **Consistent output** - Same rendering in RStudio, ggsave(), and R Markdown
✅ **No DPI issues** - Text renders correctly at any resolution
✅ **Better quality** - Superior anti-aliasing and text rendering
✅ **Simpler workflow** - One-time font installation, no per-session setup
✅ **Faster** - Better performance than showtext

# Quick Start: One-Command Setup

The easiest way to set up fonts for benviplot is to run one command:

```{r, eval=FALSE}
library(benviplot)

# One-time setup - installs Poppins and provides RStudio configuration guidance
setup_benvi_fonts()
```

This function will:

1. Download and install the Poppins font to your system
2. Check if ragg is installed (recommended for best quality)
3. Provide instructions for configuring RStudio

After running this once, you're done! The fonts will work in all R sessions.

# Detailed Installation

## Install Poppins Font

### Option 1: Automatic Installation (Recommended)

```{r, eval=FALSE}
# Install Poppins font from Google Fonts
benviplot::install_poppins()
```

This downloads and installs Poppins system-wide. You only need to do this once per computer.

### Option 2: Manual Installation

You can also install Poppins manually:

1. Go to [Google Fonts - Poppins](https://fonts.google.com/specimen/Poppins)
2. Click "Download family"
3. Extract the ZIP file
4. Install the fonts:
   - **Windows**: Right-click each .ttf file → "Install"
   - **macOS**: Double-click each .ttf file → "Install Font"
   - **Linux**: Copy .ttf files to `~/.fonts/` or use your system's font manager

After manual installation, restart R to make the fonts available.

## Install ragg (Optional but Recommended)

For the best graphics quality, install the `ragg` package:

```{r, eval=FALSE}
install.packages("ragg")
```

The ragg package provides high-quality graphics devices with excellent text rendering.

# Configure RStudio (Optional)

To use ragg as the default graphics device in RStudio:

1. Go to **Tools → Global Options → General → Graphics**
2. Set **Backend** to **AGG**
3. Click **OK** and restart RStudio

This ensures all plots in the RStudio Plots pane use ragg for rendering.

# Check Your Setup

Use `font_status()` to check your current configuration:

```{r, eval=FALSE}
font_status()
```

This will report:

- Whether Poppins is installed
- Whether ragg is available
- Recommendations for optimal setup

# Using benviplot with Fonts

## Theme Behavior

The `theme_benvi()` function automatically detects whether Poppins is installed:

- **If Poppins is installed**: Uses Poppins for all text
- **If Poppins is not installed**: Falls back to system "sans" font

```{r, eval=FALSE}
library(ggplot2)
library(benviplot)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "My Plot") +
  theme_benvi()  # Automatically uses Poppins or fallback
```

You'll see a one-time message if Poppins isn't installed, suggesting installation.

## Saving Plots

Use `ggsave_benvi()` for optimized plot saving:

```{r, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_benvi()

# Automatically uses ragg for PNG if available
ggsave_benvi("my_plot.png", p, width = 8, height = 6)

# Works with any format
ggsave_benvi("my_plot.pdf", p, width = 8, height = 6)
```

`ggsave_benvi()` automatically:

- Uses ragg device for PNG files (if ragg is installed)
- Sets DPI to 300 for high quality
- Provides consistent output across platforms

You can still use regular `ggsave()`, but `ggsave_benvi()` provides better defaults.

# Troubleshooting

## "Poppins font not found" Message

If you see this message when using `theme_benvi()`:

1. Install Poppins: `install_poppins()`
2. Restart R
3. Try again

## Font Looks Different

If the font doesn't look like Poppins:

```{r, eval=FALSE}
# Check if Poppins is actually installed
check_poppins_installed()

# If FALSE, install it
install_poppins()
```

## ragg Not Working in RStudio

1. Ensure ragg is installed: `install.packages("ragg")`
2. Configure RStudio: Tools → Global Options → General → Graphics → Backend: AGG
3. Restart RStudio

## Text Still Has Size Issues

The new system eliminates DPI issues, but if you're still seeing problems:

1. Ensure you're using benviplot v1.1.0 or later: `packageVersion("benviplot")`
2. Check that you're not loading the old showtext package
3. Use `ggsave_benvi()` instead of base `ggsave()`

# Advanced: Using Other Fonts

While benviplot is designed for Poppins, you can use any system font with the systemfonts framework:

```{r, eval=FALSE}
# List all available system fonts
systemfonts::system_fonts()

# Use a different font in theme_benvi
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_benvi() +
  theme(text = element_text(family = "Roboto"))  # Override font
```

For more advanced font management, see the [systemfonts documentation](https://systemfonts.r-lib.org/).

# Migration from v1.0.0

If you're upgrading from benviplot v1.0.0:

## What Changed

- ❌ **Removed**: `showtext` and `sysfonts` packages
- ❌ **Removed**: `import_fonts()` function (replaced with `install_poppins()`)
- ✅ **Added**: Modern systemfonts-based font management
- ✅ **Added**: New functions: `install_poppins()`, `setup_benvi_fonts()`, `font_status()`, `ggsave_benvi()`

## Migration Steps

1. Remove old showtext code from your scripts:

```{r, eval=FALSE}
# OLD (v1.0.0) - DELETE THIS
library(showtext)
showtext_auto()
import_fonts()

# NEW (v1.1.0) - One-time setup
setup_benvi_fonts()
```

2. Install Poppins system-wide (one time):

```{r, eval=FALSE}
install_poppins()
```

3. Optionally install ragg for best quality:

```{r, eval=FALSE}
install.packages("ragg")
```

4. Update your plotting code (optional):

```{r, eval=FALSE}
# OLD
ggsave("plot.png", dpi = 300)

# NEW (recommended)
ggsave_benvi("plot.png")  # Smart defaults
```

That's it! Your plots will now render consistently without any DPI issues.

# Summary

The new font system in benviplot v1.1.0 provides a dramatically better user experience:

- **One-time setup**: Install fonts to your system once
- **No DPI issues**: Consistent rendering at any resolution
- **Better quality**: Superior text rendering with ragg
- **Simpler code**: No per-session font loading required

For most users, running `setup_benvi_fonts()` once is all you need!
