---
title: "Getting Started with benviplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with benviplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

`benviplot` provides color palettes and ggplot2 helper functions tailored for EDA. The color underlying color schemes are based on Benvi, a discontinued brand of Brazilian proptech QuintoAndar. This package uses publicly available color schemes for data visualization purposes.

## Installation

The easiest way to install `benviplot` currently is directly from GitHub.

```{r, eval = FALSE}
# Install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install benviplot
remotes::install_github("viniciusoike/benviplot")
```

## Font Setup (Optional but Recommended)

`benviplot` uses the **Poppins** font family from Google Fonts. Installing Poppins is **optional** - the theme will automatically fall back to your system's default font if Poppins isn't available.

### One-Command Setup

Poppins is freely available at Google Fonts but you can also install it with `setup_benvi_fonts()`.

```{r, eval = FALSE}
library(benviplot)

# One-time setup - installs Poppins and configures graphics
setup_benvi_fonts()
```

This function additionally checks if `ragg` is installed and provides guidance for RStudio configuration. If you're running R from Positron, no additional configuration is necessary.

## Quick Start

While the package can work with base R plots, the recommended use is together with `ggplot2`.

```{r setup}
library(ggplot2)
library(benviplot)
```

The basic usage of package is centered around the `theme_benvi()` and `scale_*` functions.

```{r quick-scatter}
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(name = "Cylinders", pal_name = "qual_8") +
  labs(
    title = "Fuel Efficiency vs. Weight",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

## Viewing Color Palettes

Before using a palette, you can preview it.

```{r view-palettes}
# View a qualitative palette
benvi_palette("qual_2")

# View a sequential palette
benvi_palette("benvi_blue")

# Get hex codes as character vector
colors <- as.character(benvi_palette("qual_2"))
```

## General Examples

### Bar Chart

For simple plots it's easier to select colors using `benvi_palette` and `theme_benvi`.

```{r bar-chart}
# Sample data
sales <- data.frame(
  cities = c(
    "São Paulo",
    "Rio de Janeiro",
    "Belo Horizonte",
    "Porto Alegre",
    "Curitiba"
  ),
  revenue = c(125, 200, 150, 175, 80)
)

ggplot(sales, aes(x = cities, y = revenue)) +
  geom_col(fill = benvi_palette("benvi_blue")[3], show.legend = FALSE) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Revenue by Product",
    x = NULL,
    y = "Revenue ($1000s)"
  ) +
  theme_benvi() +
  theme(panel.grid.major.x = element_blank())
```

### Heatmap with Continuous Colors

Even though the Benvi palettes are discrete, applying them to a continuous variable still works.

```{r heatmap}
# Sample data - Texas housing
housing <- subset(txhousing, city %in% c("Austin", "Houston", "Dallas"))

ggplot(housing, aes(x = date, y = city, fill = log10(median))) +
  geom_tile() +
  scale_fill_benvi_c(
      pal_name = "benvi_blue",
      name = "Median\nPrice ($)",
      direction = -1) +
  labs(
    title = "Median House Prices in Texas Cities",
    x = "Year",
    y = NULL
  ) +
  theme_benvi()
```

## Using Plot Helper Functions

`benviplot` includes convenience functions for common plot types. These save typing during exploratory data analysis. These functions come with several changes to default `ggplot2` values. The first example shows a simple line plot using the `economics` dataset.

```{r helper-line}
plot_line(economics, x = date, y = uempmed)
```

The plot generated above is essentially a wrapper around other `ggplot2` functions as arguments.

```{r, eval = FALSE}
# Makes the same plot as the one above
ggplot(economics, aes(x = date, y = uempmed)) +
  geom_line(lwd = 1, color = benvi_palette("benvi_blue", 1)) +
  geom_hline(yintercept = 0) +
  labs(x = NULL) +
  theme_benvi()
```

The helper functions include several convenient arguments such as `text` that add text labels on bar plots.

```{r helper-column}
plot_column(sales, x = cities, y = revenue, text = TRUE)
```

These functions also support using aesthetics as grouping variables under a generic `variable` alias.

```{r helper-scatter}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  variable = as.factor(cyl),
  palette = "qual_8",
  scale_name = "Cylinders"
)
```

Finally, helper functions return `ggplot2` objects, so you can add layers normally.

```{r customization}
plot_line(economics, x = date, y = unemploy / 1000) +
  geom_smooth(se = FALSE, color = benvi_palette("oranges")[3]) +
  labs(
    title = "US Unemployment with Smoothed Trend",
    subtitle = "Unemployment figures in millions",
    x = "Year",
    y = "Unemployed (millions)",
    caption = "Data: economics dataset"
  )
```

### Can I use benvi colors in base R plots?

You can use benvi colors normally in base R plots. However, replicating `ggplot2` functionality might be challenging.

```{r}
colors <- as.character(benvi_palette("purples"))

plot(
  mtcars$wt,
  mtcars$mpg,
  col = colors[mtcars$cyl / 2 - 1],
  pch = 19,
  cex = 1.5,
  xlab = "Weight",
  ylab = "MPG",
  main = "Using benvi colors in base R"
)
```

## Next Steps

1.  **Explore all palettes**: See `vignette("color-palettes")` for complete gallery.
2.  **Learn plot functions**: Check `vignette("plot-functions")` for detailed examples.
3.  **Customize themes**: Read `vignette("themes-and-styling")` for advanced styling.

## Getting Help

-   **Function documentation**: `?benvi_palette`, `?scale_color_benvi_d`, etc.
-   **Package website**: https://viniciusoike.github.io/benviplot/
-   **Report issues**: https://github.com/viniciusoike/benviplot/issues
-   **Ask questions**: Open a GitHub discussion.

Happy plotting!
