---
title: "Getting Started with benviplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with benviplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

`benviplot` provides color palettes and ggplot2 helper functions for exploratory
data analysis. The color schemes are based on Benvi, a brand of the Brazilian
proptech QuintoAndar.[^1] The package includes a custom ggplot2 theme, discrete
and continuous color scales, and wrappers for common chart types.

[^1]: The color information is publicly available. This project is not
    affiliated with QuintoAndar.

### Installation

```{r, eval = FALSE}
# Install remotes if needed
install.packages("remotes")

remotes::install_github("viniciusoike/benviplot")
```

### Optional font setup

`benviplot` bundles the Poppins font family. When both `systemfonts` and `ragg`
are available, the package registers Poppins and `theme_benvi()` uses it by
default. Otherwise, the theme uses the system sans-serif font.

Install `ragg` to render Poppins in PNG files. Base PDF and PostScript devices
cannot use fonts registered through `systemfonts`; use `base_family = "sans"`
with those devices.

```{r, eval = FALSE}
install.packages("ragg")
```

Check whether Poppins and `ragg` are available with `font_status()`.

```{r, eval = FALSE}
font_status()
```

## Quick start

```{r setup, include = FALSE}
library(ggplot2)
library(benviplot)
```

```{r load}
library(ggplot2)
library(benviplot)
```

Combine `theme_benvi()` with the `scale_*_benvi_*()` functions to style a
ggplot2 chart. In the example below, `scale_fill_benvi_d()` maps the discrete
cylinder variable to the `"qual_8"` qualitative palette. `theme_benvi()` sets
the remaining theme elements.

```{r quick-scatter}
ggplot(mtcars, aes(x = wt, y = mpg, fill = as.factor(cyl))) +
  geom_point(shape = 21, size = 3, color = "#000000") +
  scale_fill_benvi_d(name = "Cylinders", pal_name = "qual_8") +
  labs(
    title = "Fuel Efficiency vs. Weight",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

## Color palettes

The package organizes its palettes into theme, sequential, qualitative,
diverging, city-specific, and brand families. Choose a family based on the
variable and the role of color in the chart.

### Browsing palettes

`show_palettes()` displays the available palettes. Calling it without arguments
shows every palette; pass a type name to narrow the output.

```{r show-all, fig.asp = 2.5, eval = FALSE}
show_palettes()
```

Filter the display with `"theme"`, `"sequential"`, `"qualitative"`,
`"diverging"`, `"city"`, or `"brand"`.

```{r show-seq}
show_palettes("sequential")
```

### Accessing palette colors

`benvi_palette()` returns a `palette` object backed by hexadecimal color values.
Printing the object draws color swatches on the active graphics device.

```{r view-palettes}
# Preview a palette
benvi_palette("qual_2")
```

To pass the colors to another package, coerce the result to a plain character
vector.

```{r}
# Get hex codes as a plain character vector
as.character(benvi_palette("benvi_blue"))
```

Discrete palettes contain between 4 and 9 fixed colors. Set
`type = "continuous"` to interpolate more colors along the palette gradient.

```{r continuous}
benvi_palette("seq_greens", n = 20, type = "continuous")
```

## Using ggplot2 scales

The scale functions follow the ggplot2 naming pattern
`scale_{aesthetic}_benvi_{d|c}()`. The suffix `d` denotes a discrete scale, and
`c` denotes a continuous scale. Both `color` and `fill` variants are available;
`colour` spellings are aliases of the `color` functions.

### Discrete scales

Discrete scales map categorical variables to fixed colors. Qualitative palettes
usually work best for this purpose, but `pal_name` accepts any package palette.

```{r discrete-scale}
iqaiw_total <- subset(iqaiw, rooms == "Total")

ggplot(iqaiw_total, aes(x = date, y = index, color = name_muni)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 100) +
  scale_color_benvi_d("rio_qual", name = NULL) +
  labs(
    title = "IQAIW Rental Index by City",
    x = NULL,
    y = "Index (base = 100)"
  ) +
  theme_benvi()
```

### Continuous scales

Continuous scales interpolate colors from any package palette. Sequential
palettes usually work best for ordered numeric values. Use `direction = -1` to
reverse the palette.

```{r continuous-scale}
iqaiw_total <- subset(iqaiw_total, !is.na(acum12m))

ggplot(iqaiw_total, aes(x = date, y = name_muni, fill = acum12m * 100)) +
  geom_tile(height = 0.6, color = "#ffffff") +
  scale_fill_benvi_c(
    pal_name = "benvi_blue",
    name = "YoY Change (%)",
    direction = -1
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand = expansion(0)
  ) +
  labs(x = NULL, y = NULL) +
  theme_benvi() +
  theme(
    legend.title = element_text(hjust = 0.5, vjust = 0.75),
    axis.text = element_text(size = 12),
    panel.grid = element_blank()
  )
```

## Plot helper functions

`benviplot` includes wrappers for common chart types. These functions accept a
data frame and unquoted column names, apply `theme_benvi()`, and return a
`ggplot` object. Build charts directly with ggplot2 when you need finer control.

### Line chart

`plot_line()` draws a single-series line chart. Pass `color` to map a grouping
variable and draw multiple lines with a legend.

```{r helper-line}
spo_index <- subset(iqaiw, name_muni == "São Paulo" & rooms == "Total")
plot_line(spo_index, x = date, y = index)
```

### Bar chart

`plot_column()` creates a vertical bar chart. Setting `text = TRUE` adds value
labels above each bar. Set `text_inside = TRUE` to place the labels inside the
bars; this option requires the `ggfittext` package.

```{r helper-column}
latest_sales <- subset(
  sales_report,
  name_muni == "Belo Horizonte" & date == max(date)
)

plot_column(latest_sales, x = name_zone, y = price_m2, text = TRUE)
```

### Scatter plot

`plot_scatter()` maps `x` and `y` to a scatter plot. Pass a variable to `color`
to apply a Benvi palette to the points. Set `fit = TRUE` to add a fitted line.

```{r helper-scatter}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  color = as.factor(cyl),
  pal_name = "qual_5",
  scale_name = "Cylinders",
  fit = TRUE
)
```

### Area chart

`plot_area()` draws a single area or maps a grouping variable to `fill` to draw
stacked areas.

```{r helper-area}
room_index <- subset(
  iqaiw,
  name_muni == "São Paulo" & rooms %in% c("1", "2", "3")
)

plot_area(room_index, x = date, y = index, fill = rooms)
```

### Histogram

`plot_histogram()` selects the bin width with the Freedman–Diaconis rule by
default. Set `bins` to choose the number of bins directly.

```{r helper-histogram}
plot_histogram(mtcars, x = mpg)
```

### Adding layers

Because the helpers return `ggplot` objects, you can add layers, scales, and
theme adjustments with the usual ggplot2 syntax.

```{r customization}
plot_line(spo_index, x = date, y = index) +
  geom_smooth(se = FALSE, color = benvi_palette("oranges")[3]) +
  labs(
    title = "Rental Price Index in São Paulo",
    subtitle = "Smoothed trend",
    x = NULL,
    y = "Index (base = 100)",
    caption = "Source: IQAIW"
  )
```

## Base R

The palettes are not tied to ggplot2. Convert a palette to a character vector
to use its hexadecimal color values with base R graphics, lattice, or another
plotting system.

```{r base-r}
colors <- as.character(benvi_palette("purples"))

plot(
  mtcars$wt,
  mtcars$mpg,
  col = colors[mtcars$cyl / 2 - 1],
  pch = 19,
  cex = 1.5,
  xlab = "Weight (1000 lbs)",
  ylab = "Miles per Gallon",
  main = "Using Benvi colors in base R"
)
```

## Getting help

- Open function documentation with `?benvi_palette`, `?theme_benvi`, or
  `?scale_color_benvi_d`.
- Browse the [package website](https://viniciusoike.github.io/benviplot/).
- Report problems in the
  [GitHub issue tracker](https://github.com/viniciusoike/benviplot/issues).
