---
title: "Getting Started with benviplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with benviplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Welcome to `benviplot`! This package provides a comprehensive set of tools for creating beautiful, publication-ready visualizations in R using ggplot2.

### What is benviplot?

`benviplot` offers:

- **Curated Color Palettes**: Professional color schemes optimized for data visualization
- **ggplot2 Integration**: Seamless scales for both discrete and continuous data
- **Plot Helpers**: Convenience functions for common chart types
- **Custom Theme**: Clean, modern theme with professional typography

### Disclaimer

**IMPORTANT**: This is an unofficial, independent project created by Vinicius Oike and is **NOT** affiliated with, endorsed by, or connected to QuintoAndar in any way. Benvi was a former brand of QuintoAndar that was discontinued in 2024. This package uses publicly available color schemes from that period for data visualization purposes.

## Installation

Install `benviplot` from GitHub:

```{r, eval = FALSE}
# Install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install benviplot
remotes::install_github("viniciusoike/benviplot")
```

### System Requirements

- **R**: Version 4.1.0 or higher
- **ggplot2**: Version 4.0.0 or higher
- **Internet connection**: Required for initial font download

## Font Setup

`benviplot` uses the **Poppins** font family from Google Fonts to create professional-looking plots. The package will attempt to download and configure this font automatically on first use.

### Automatic Font Import

Simply load the package - fonts will be set up automatically:

```{r, eval = FALSE}
library(benviplot)
#> Importing Poppins font from Google Fonts...
#> ✔ Font imported successfully
```

### Manual Font Import

If you encounter font issues, manually import fonts:

```{r, eval = FALSE}
benviplot::import_fonts()
```

### Troubleshooting Font Issues

If fonts don't display correctly:

1. **Check internet connection**: Font download requires internet access
2. **Restart R session**: Sometimes fonts need a fresh session to load
3. **Use system fonts as fallback**: Modify theme to use system fonts

```{r, eval = FALSE}
# Example: Override with system font
theme_benvi(base_family = "sans")
```

## Quick Start

Let's create your first plot with benviplot!

```{r setup}
library(ggplot2)
library(benviplot)

# Use built-in mtcars dataset
head(mtcars)
```

### A Simple Scatter Plot

```{r quick-scatter}
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "Qual2", name = "Cylinders") +
  labs(
    title = "Fuel Efficiency vs. Weight",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

That's it! Just three key components:

1. `scale_color_benvi_d()`: Applies benvi discrete color palette
2. `theme_benvi()`: Applies benvi visual theme
3. `palette = "Qual2"`: Selects specific color palette

## Viewing Color Palettes

Before using a palette, you can preview it:

```{r view-palettes}
# View a qualitative palette
benvi_palette("Qual2")

# View a sequential palette
benvi_palette("Seq3")

# View a set palette
benvi_palette("Set5")
```

### Getting Hex Color Values

Extract hex codes for use outside ggplot2:

```{r hex-colors}
# Get hex codes as character vector
colors <- as.character(benvi_palette("Qual2"))
colors

# Use specific colors
my_blue <- benvi_palette("Qual2")[1]
my_blue
```

## Basic Examples

### Bar Chart with Discrete Colors

```{r bar-chart}
# Sample data
sales <- data.frame(
  product = c("A", "B", "C", "D", "E"),
  revenue = c(120, 200, 150, 180, 220)
)

ggplot(sales, aes(x = product, y = revenue, fill = product)) +
  geom_col(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "Set3") +
  labs(
    title = "Revenue by Product",
    x = "Product",
    y = "Revenue ($1000s)"
  ) +
  theme_benvi()
```

### Line Chart with Groups

```{r line-chart}
# Use built-in economics dataset
ggplot(economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = benvi_palette("Set5")[1], linewidth = 1) +
  labs(
    title = "US Unemployment Over Time",
    x = "Year",
    y = "Unemployed (millions)"
  ) +
  theme_benvi()
```

### Heatmap with Continuous Colors

```{r heatmap}
# Sample data - Texas housing
housing <- subset(txhousing, city %in% c("Austin", "Houston", "Dallas"))

ggplot(housing, aes(x = date, y = city, fill = median)) +
  geom_tile() +
  scale_fill_benvi_c(pal_name = "Seq5", name = "Median\nPrice ($)") +
  labs(
    title = "Median House Prices in Texas Cities",
    x = "Year",
    y = NULL
  ) +
  theme_benvi()
```

## Using Plot Helper Functions

`benviplot` includes convenience functions for common plot types. These save typing during exploratory data analysis.

### Simple Line Plot

```{r helper-line}
plot_line(economics, x = date, y = uempmed)
```

### Column Plot with Labels

```{r helper-column}
plot_column(sales, x = product, y = revenue, text = TRUE)
```

### Scatter Plot with Grouping

```{r helper-scatter}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  variable = as.factor(cyl),
  palette = "Qual5",
  scale_name = "Cylinders"
)
```

### Adding Trend Lines

```{r scatter-trend}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  variable = as.factor(cyl),
  fit = TRUE,
  fit_method = "lm",
  palette = "Qual5",
  scale_name = "Cylinders"
) +
  labs(
    title = "Fuel Efficiency by Weight and Cylinders",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )
```

## Customizing Your Plots

Helper functions return ggplot2 objects, so you can add layers:

```{r customization}
plot_line(economics, x = date, y = unemploy / 1000) +
  geom_smooth(se = FALSE, color = benvi_palette("Set1")[2]) +
  labs(
    title = "US Unemployment with Smoothed Trend",
    subtitle = "Unemployment figures in millions",
    x = "Year",
    y = "Unemployed (millions)",
    caption = "Data: economics dataset"
  ) +
  theme_benvi()
```

## Understanding Palette Types

`benviplot` organizes palettes by use case:

### Qualitative Palettes (Qual1-Qual9)

**Use for**: Categorical data without inherent order (e.g., product types, regions)

- 8 distinct colors
- Optimized for visual differentiation
- Examples: `Qual1`, `Qual2`, ..., `Qual9`

```{r qual-example, eval = FALSE}
benvi_palette("Qual3")
```

### Sequential Palettes (Seq0-Seq7)

**Use for**: Continuous data with ordered progression (e.g., temperature, percentages)

- 9-color gradients
- Light to dark progression
- Examples: `Seq0`, `Seq1`, ..., `Seq7`

```{r seq-example, eval = FALSE}
benvi_palette("Seq2")
```

### Set Palettes (Set0-Set7)

**Use for**: Small categorical datasets (2-4 categories)

- 4 carefully chosen colors
- Good contrast and balance
- Examples: `Set0`, `Set1`, ..., `Set7`

```{r set-example, eval = FALSE}
benvi_palette("Set4")
```

### City-Specific Palettes

Themed palettes for Brazilian cities:

- `spo_qual`, `spo_seq`, `spo_div`: São Paulo
- `rio_qual`, `rio_seq`, `rio_div`: Rio de Janeiro
- `bhe_qual`, `bhe_seq`, `bhe_div`: Belo Horizonte

```{r city-example, eval = FALSE}
benvi_palette("rio_qual")
```

## Next Steps

Now that you're familiar with the basics:

1. **Explore all palettes**: See `vignette("color-palettes")` for complete gallery
2. **Learn plot functions**: Check `vignette("plot-functions")` for detailed examples
3. **Customize themes**: Read `vignette("themes-and-styling")` for advanced styling

## Common Questions

### How do I choose the right palette?

- **Categorical data** (no order): Use `Qual*` palettes
- **Continuous/ordered data**: Use `Seq*` palettes
- **Small categories (≤4)**: Use `Set*` palettes
- **Geographic data**: Consider city-specific palettes

### Can I reverse a palette?

Yes! Use `direction = -1`:

```{r reverse-palette}
# Normal direction
benvi_palette("Seq3")

# Reversed
benvi_palette("Seq3", direction = -1)
```

### How do I use fewer colors than available?

Specify the number with `n`:

```{r subset-colors}
# Get only 3 colors from an 8-color palette
benvi_palette("Qual5", n = 3)
```

### Can I use benvi colors in base R plots?

Yes! Get hex codes and use them:

```{r base-plot}
colors <- as.character(benvi_palette("Set5"))

plot(
  mtcars$wt,
  mtcars$mpg,
  col = colors[mtcars$cyl / 2 - 1],
  pch = 19,
  cex = 1.5,
  xlab = "Weight",
  ylab = "MPG",
  main = "Using benvi colors in base R"
)
```

## Getting Help

- **Function documentation**: `?benvi_palette`, `?scale_color_benvi_d`, etc.
- **Package website**: https://viniciusoike.github.io/benviplot/
- **Report issues**: https://github.com/viniciusoike/benviplot/issues
- **Ask questions**: Open a GitHub discussion

## Summary

In this vignette you learned how to:

- ✓ Install and set up benviplot
- ✓ Configure Poppins fonts
- ✓ View and select color palettes
- ✓ Create plots with benvi scales and themes
- ✓ Use plot helper functions
- ✓ Choose appropriate palettes for your data

Happy plotting!
