---
title: "Getting Started with benviplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with benviplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

`benviplot` provides color palettes and ggplot2 helper functions for exploratory data analysis. The color schemes are based on Benvi, a discontinued brand of the Brazilian proptech QuintoAndar[^1]. The package ships with a custom ggplot2 theme, a family of discrete and continuous color scales, and convenience wrappers for common chart types.

[^1]: All color information used here is publicly available and this project is not affiliated with QuintoAndar.

### Installation

```{r, eval = FALSE}
# Install remotes if needed
install.packages("remotes")

remotes::install_github("viniciusoike/benviplot")
```

### Font Setup (Optional)

`benviplot` bundles the Poppins font family and registers it automatically when
the package loads (requires the `systemfonts` package). By default,
`theme_benvi()` uses the system sans-serif font so nothing breaks on machines
that lack `systemfonts` or use a base PDF/PostScript device.

To activate Poppins, set the option once per session:

```{r, eval = FALSE}
options(theme_benvi.font_family = "Poppins")
```

To make this permanent across sessions, add the same line to your `~/.Rprofile`.
You can open it with `usethis::edit_r_profile()`.

For the best rendering quality with custom fonts, install the `ragg` package.
When `ragg` is set as the graphics device (the default in RStudio and Positron),
Poppins renders correctly in all output formats.

```{r, eval = FALSE}
install.packages("ragg")
```

You can check your full font and device status at any time with:

```{r, eval = FALSE}
font_status()
```

## Quick Start

```{r setup, include = FALSE}
library(ggplot2)
library(benviplot)
options(theme_benvi.font_family = "Poppins")
```

```{r load}
library(ggplot2)
library(benviplot)
```

The core of the package is `theme_benvi()` combined with the `scale_*_benvi_*()` functions. Together, they give any ggplot2 chart a consistent look with minimal effort. In the example below, `scale_fill_benvi_d()` maps the discrete cylinder variable to the `"qual_8"` qualitative palette, while `theme_benvi()` handles the rest of the styling.

```{r quick-scatter}
ggplot(mtcars, aes(x = wt, y = mpg, fill = as.factor(cyl))) +
  geom_point(shape = 21, size = 3, color = "#000000") +
  scale_fill_benvi_d(name = "Cylinders", pal_name = "qual_8") +
  labs(
    title = "Fuel Efficiency vs. Weight",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

## Color Palettes

The package organizes its palettes into several families: theme, sequential, qualitative, diverging, city-specific, and brand. Each family serves a different purpose, and choosing the right one depends on the nature of your variable.

### Browsing palettes

`show_palettes()` gives a visual overview of all available palettes. Calling it without arguments displays every palette in the package; pass a type name to narrow the output.

```{r show-all, fig.asp = 2.5, eval = FALSE}
show_palettes()
```

You can filter by type: `"theme"`, `"sequential"`, `"qualitative"`, `"city"`, or `"brand"`.

```{r show-seq}
show_palettes("sequential")
```

### Accessing palette colors

`benvi_palette()` returns a named vector of hex codes. Printing it renders a color swatch in the console, which is useful for quick visual comparison.

```{r view-palettes}
# Preview a palette
benvi_palette("qual_2")
```

To use the colors outside of ggplot2 (e.g. in base R or as input to another package), coerce to a plain character vector.

```{r}
# Get hex codes as a plain character vector
as.character(benvi_palette("benvi_blue"))
```

Discrete palettes contain between 4 and 9 fixed colors. When you need more granularity, set `type = "continuous"` to interpolate an arbitrary number of colors along the palette gradient.

```{r continuous}
benvi_palette("seq_greens", n = 20, type = "continuous")
```

## Using ggplot2 Scales

The scale functions follow the standard ggplot2 naming convention: `scale_{aesthetic}_benvi_{d|c}()`, where `d` is for discrete variables and `c` is for continuous ones. Both `color` and `fill` variants are available, and `colour` spellings work as expected.

### Discrete scales

Discrete scales map categorical variables to a qualitative or city-specific palette. The `pal_name` argument (or the first positional argument) selects which palette to use.

```{r discrete-scale}
iqaiw_total <- subset(iqaiw, rooms == "Total")

ggplot(iqaiw_total, aes(x = date, y = index, color = name_muni)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 100) +
  scale_color_benvi_d("rio_qual", name = NULL) +
  labs(
    title = "IQAIW Rental Index by City",
    x = NULL,
    y = "Index (base = 100)"
  ) +
  theme_benvi()
```

### Continuous scales

Continuous scales interpolate across a sequential or brand palette. They are well suited for heatmaps, choropleths, and any visualization where a numeric variable needs a smooth color gradient. Use `direction = -1` to reverse the palette.

```{r continuous-scale}
iqaiw_total <- subset(iqaiw_total, !is.na(acum12m))

ggplot(iqaiw_total, aes(x = date, y = name_muni, fill = acum12m * 100)) +
  geom_tile(height = 0.6, color = "#ffffff") +
  scale_fill_benvi_c(
    pal_name = "benvi_blue",
    name = "YoY Change (%)",
    direction = -1
  ) +
  scale_x_continuous(
    breaks = seq(2023, 2025, 1),
    expand = expansion(0)
  ) +
  labs(x = NULL, y = NULL) +
  theme_benvi() +
  theme(
    legend.title = element_text(hjust = 0.5, vjust = 0.75),
    axis.text = element_text(size = 12),
    panel.grid = element_blank()
  )
```

## Plot Helper Functions

`benviplot` includes convenience wrappers for common chart types. These functions accept a data frame and column names via tidy evaluation, apply `theme_benvi()` automatically, and return a standard `ggplot2` object. They are designed for quick exploratory work; for publication-quality figures, building the plot from scratch gives you full control.

### Line chart

`plot_line()` draws a single-series line chart. Pass `color` to map a grouping variable and get multiple lines with an automatic legend.

```{r helper-line}
plot_line(economics, x = date, y = uempmed)
```

### Bar chart

`plot_column()` creates a vertical bar chart. Setting `text = TRUE` adds value labels above each bar; for bars with enough height, `text_inside = TRUE` places labels inside the bars instead (requires the `ggfittext` package).

```{r helper-column}
sales <- data.frame(
  cities = c(
    "São Paulo",
    "Rio de Janeiro",
    "Belo Horizonte",
    "Porto Alegre",
    "Curitiba"
  ),
  revenue = c(125, 200, 150, 175, 80)
)

plot_column(sales, x = cities, y = revenue, text = TRUE)
```

### Scatter plot

`plot_scatter()` maps `x` and `y` to a scatter plot. A `color` argument maps a grouping variable to the point fill using a Benvi palette. Set `smooth = TRUE` to overlay a regression line.

```{r helper-scatter}
plot_scatter(
  mtcars,
  x = wt,
  y = mpg,
  color = as.factor(cyl),
  palette = "qual_5",
  scale_name = "Cylinders"
)
```

### Adding layers

Because all helpers return `ggplot2` objects, you can extend them with additional layers, scales, or theme adjustments as you normally would.

```{r customization}
plot_line(economics, x = date, y = unemploy / 1000) +
  geom_smooth(se = FALSE, color = benvi_palette("oranges")[3]) +
  labs(
    title = "US Unemployment with Smoothed Trend",
    subtitle = "Unemployment figures in millions",
    x = NULL,
    y = "Unemployed (millions)",
    caption = "Data: economics dataset"
  )
```

## Base R

The palettes are not tied to ggplot2. Since `benvi_palette()` returns plain hex codes, you can use them anywhere that accepts color strings, including base R graphics, `lattice`, or any other plotting system.

```{r base-r}
colors <- as.character(benvi_palette("purples"))

plot(
  mtcars$wt,
  mtcars$mpg,
  col = colors[mtcars$cyl / 2 - 1],
  pch = 19,
  cex = 1.5,
  xlab = "Weight (1000 lbs)",
  ylab = "Miles per Gallon",
  main = "Using Benvi colors in base R"
)
```

## Getting Help

-   **Function documentation**: `?benvi_palette`, `?theme_benvi`, `?scale_color_benvi_d`
-   **Package website**: https://viniciusoike.github.io/benviplot/
-   **Report issues**: https://github.com/viniciusoike/benviplot/issues
