---
title: "Themes and Styling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Themes and Styling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ggplot2)
library(benviplot)
```

## Introduction

Visual consistency is key to professional data visualization. This vignette shows you how to use and customize `theme_benvi()` to create polished, publication-ready graphics.

## The theme_benvi() Theme

`theme_benvi()` provides a clean, modern aesthetic designed for clarity and readability:

- **Clean background**: White with minimal gridlines
- **Professional typography**: Poppins font family
- **Smart defaults**: Legend on top, appropriate text sizes
- **Flexible**: Easy to customize

### Basic Usage

Simply add `theme_benvi()` to any ggplot:

```{r basic-theme}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("Set5")[1], size = 3) +
  labs(
    title = "Fuel Efficiency vs. Weight",
    subtitle = "Using theme_benvi()",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_benvi()
```

### Comparing with Default Theme

```{r compare-themes, fig.height = 8}
# Create base plot
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "Qual5", name = "Cylinders") +
  labs(
    title = "Fuel Efficiency Analysis",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )

# Default ggplot2 theme
p1 <- p + labs(subtitle = "Default theme_gray()") + theme_gray()

# benvi theme
p2 <- p + labs(subtitle = "theme_benvi()") + theme_benvi()

# Display side by side (requires patchwork or gridExtra)
# For vignette, we'll show separately
p1
p2
```

## Theme Components

### Typography

`theme_benvi()` uses the **Poppins** font with carefully chosen sizes:

- **Title**: 12pt, black
- **Subtitle**: 8pt, dark gray
- **Body text**: 10pt, dark gray
- **Caption**: 6pt, dark gray

```{r typography}
ggplot(economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = benvi_palette("Set1")[1], linewidth = 1) +
  labs(
    title = "Title: Clear and Prominent",
    subtitle = "Subtitle: Provides additional context",
    x = "X-axis label",
    y = "Y-axis label",
    caption = "Caption: Source information or notes"
  ) +
  theme_benvi()
```

### Grid Lines

Only major grid lines are shown for cleaner appearance:

```{r grid-comparison}
# With theme_benvi (major grid only)
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = benvi_palette("Set3")[1]) +
  labs(title = "Clean Grid: Major Lines Only") +
  theme_benvi()
```

### Legend Position

Default legend position is on top for better readability:

```{r legend-top}
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "Set5", name = "Cylinders") +
  labs(title = "Legend on Top (Default)") +
  theme_benvi()
```

## Font Management

### Automatic Font Loading

`benviplot` automatically downloads and configures the Poppins font on first use:

```{r font-auto, eval = FALSE}
library(benviplot)
# Font loads automatically
```

### Manual Font Import

If needed, manually import fonts:

```{r font-manual, eval = FALSE}
benviplot::import_fonts()
```

### Troubleshooting Fonts

If fonts don't display:

**1. Check Internet Connection**

Font download requires internet access.

**2. Restart R Session**

```{r restart, eval = FALSE}
# Restart R, then:
library(benviplot)
```

**3. Verify Font Installation**

```{r verify-font, eval = FALSE}
# Check available fonts
sysfonts::font_families()
```

**4. Use Fallback Font**

Override with system font if needed:

```{r fallback, eval = FALSE}
# Not directly supported, but you can modify the theme
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_benvi() +
  theme(text = element_text(family = "sans"))
```

## Customizing theme_benvi()

`theme_benvi()` is a starting point. Customize it with additional `theme()` calls:

### Change Legend Position

```{r custom-legend-pos}
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 3) +
  scale_color_benvi_d(pal_name = "Set7", name = "Cylinders") +
  labs(title = "Legend on Right") +
  theme_benvi() +
  theme(legend.position = "right")
```

### Adjust Text Sizes

```{r custom-text-size}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("Set2")[1], size = 3) +
  labs(
    title = "Larger Text for Presentations",
    x = "Weight",
    y = "MPG"
  ) +
  theme_benvi() +
  theme(
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

### Rotate Axis Labels

```{r custom-axis-rotation}
products <- data.frame(
  product = c("Product A", "Product B", "Product C", "Product D"),
  sales = c(120, 200, 150, 180)
)

ggplot(products, aes(x = product, y = sales)) +
  geom_col(fill = benvi_palette("Set4")[1]) +
  labs(title = "Rotated X-Axis Labels") +
  theme_benvi() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Remove Grid Lines

```{r custom-no-grid}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("rio_qual")[1], size = 3) +
  labs(title = "No Grid Lines") +
  theme_benvi() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Add Background Color

```{r custom-background}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_area(fill = benvi_palette("Seq5")[5], alpha = 0.7) +
  labs(title = "Custom Background Color") +
  theme_benvi() +
  theme(
    plot.background = element_rect(fill = "#f5f5f5"),
    panel.background = element_rect(fill = "#ffffff")
  )
```

## Creating Publication-Ready Plots

### Elements of a Publication Plot

1. **Clear title and subtitle**
2. **Labeled axes with units**
3. **Data source in caption**
4. **Appropriate color palette**
5. **Legend (if needed)**
6. **Clean, minimal design**

```{r publication-example}
# Sample data
housing_subset <- txhousing |>
  dplyr::filter(city %in% c("Austin", "Houston", "Dallas"))

# Publication-ready plot
ggplot(housing_subset, aes(x = date, y = median / 1000, color = city)) +
  geom_line(linewidth = 1.2) +
  scale_color_benvi_d(
    pal_name = "Set5",
    name = "City"
  ) +
  labs(
    title = "Median Home Prices in Major Texas Cities",
    subtitle = "2000-2015",
    x = "Year",
    y = "Median Price ($1000s)",
    caption = "Source: Texas A&M Real Estate Center"
  ) +
  theme_benvi() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
```

### Faceted Plot

```{r publication-facet}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("Qual5")[1], size = 2) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = benvi_palette("Qual5")[2]
  ) +
  facet_wrap(~ cyl, labeller = label_both) +
  labs(
    title = "Fuel Efficiency by Cylinder Count",
    subtitle = "Linear regression fits shown",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    caption = "Data: mtcars dataset"
  ) +
  theme_benvi() +
  theme(
    strip.text = element_text(size = 10, face = "bold")
  )
```

## Advanced Customization

### Creating a Custom Theme

Build on `theme_benvi()` to create your own theme:

```{r custom-theme}
# Define custom theme
theme_presentation <- function() {
  theme_benvi() +
    theme(
      # Larger text for visibility
      text = element_text(size = 14),
      plot.title = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(size = 14),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12),

      # Legend
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      legend.position = "bottom",

      # More spacing
      plot.margin = margin(20, 20, 20, 20)
    )
}

# Use custom theme
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point(size = 4) +
  scale_color_benvi_d(pal_name = "Set5", name = "Cylinders") +
  labs(
    title = "Presentation-Sized Plot",
    subtitle = "Larger text for readability"
  ) +
  theme_presentation()
```

### Theme for Reports

```{r theme-report}
theme_report <- function() {
  theme_benvi() +
    theme(
      # Smaller, compact
      text = element_text(size = 9),
      plot.title = element_text(size = 11),
      plot.subtitle = element_text(size = 9),

      # No legend title to save space
      legend.title = element_blank(),
      legend.position = "top",

      # Tight margins
      plot.margin = margin(5, 5, 5, 5)
    )
}

# Use in report
ggplot(economics, aes(x = date, y = uempmed)) +
  geom_line(color = benvi_palette("Set1")[1]) +
  labs(
    title = "Median Unemployment Duration",
    caption = "Source: FRED"
  ) +
  theme_report()
```

## Combining Themes with Scales

Always use benvi scales with benvi theme for consistency:

```{r theme-scale-combo}
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "Set6") +  # benvi scale
  labs(
    title = "MPG Distribution by Cylinders",
    x = "Number of Cylinders",
    y = "Miles per Gallon"
  ) +
  theme_benvi()  # benvi theme
```

## Saving Plots

### For Print

```{r save-print, eval = FALSE}
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = benvi_palette("Set5")[1], size = 3) +
  labs(title = "Fuel Efficiency") +
  theme_benvi()

# Save high-resolution PNG
ggsave("plot.png", p, width = 7, height = 5, dpi = 300)

# Save PDF (vector graphics)
ggsave("plot.pdf", p, width = 7, height = 5)
```

### For Presentations

```{r save-presentation, eval = FALSE}
# Larger size for slides
ggsave("slide.png", p, width = 10, height = 6, dpi = 150)
```

### For Web

```{r save-web, eval = FALSE}
# Optimized for web
ggsave("web_plot.png", p, width = 8, height = 5, dpi = 96)
```

## Accessibility Tips

### Color Considerations

```{r accessibility-color}
# Use patterns in addition to colors
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_violin(show.legend = FALSE) +
  scale_fill_benvi_d(pal_name = "Set5") +
  labs(
    title = "Distribution Shape Helps Convey Information",
    subtitle = "Don't rely solely on color",
    x = "Cylinders"
  ) +
  theme_benvi()
```

### Sufficient Contrast

Ensure text is readable:

```{r accessibility-contrast}
# Good contrast between text and background
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = benvi_palette("Set1")[1]) +
  labs(
    title = "Clear, High-Contrast Text",
    x = "Weight",
    y = "MPG"
  ) +
  theme_benvi() +
  theme(
    plot.background = element_rect(fill = "white"),
    text = element_text(color = "black")
  )
```

### Alt Text for Reports

When including plots in documents, always add alt text:

```{r alt-text, eval = FALSE}
# In RMarkdown, use fig.alt
# ```{r, fig.alt="Scatter plot showing negative correlation between weight and MPG"}
# plot here
# ```
```

## Common Styling Patterns

### Minimal Plot

```{r pattern-minimal}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = benvi_palette("Set1")[1], linewidth = 1) +
  theme_benvi() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  ) +
  labs(title = "US Unemployment")
```

### Data-Focused (No Decorations)

```{r pattern-data-focused}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = benvi_palette("rio_qual")[1]) +
  theme_benvi() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "gray30"),
    axis.ticks = element_line(color = "gray30")
  )
```

### Highlighted Plot

```{r pattern-highlight}
# Emphasize one series
mtcars_highlight <- mtcars
mtcars_highlight$highlight <- ifelse(mtcars$cyl == 6, "Six", "Other")

ggplot(mtcars_highlight, aes(x = wt, y = mpg, color = highlight)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("Six" = benvi_palette("Set5")[1], "Other" = "gray70")
  ) +
  labs(
    title = "Highlighting 6-Cylinder Vehicles",
    subtitle = "Others shown in gray",
    color = NULL
  ) +
  theme_benvi()
```

## Summary

In this vignette you learned:

- ✓ How to use `theme_benvi()` for professional plots
- ✓ Font management and troubleshooting
- ✓ Customizing theme elements
- ✓ Creating publication-ready graphics
- ✓ Building custom themes
- ✓ Accessibility considerations
- ✓ Common styling patterns

## Best Practices Checklist

- [ ] Use `theme_benvi()` for consistency
- [ ] Always label axes with units
- [ ] Add informative titles and subtitles
- [ ] Include data sources in captions
- [ ] Choose appropriate palettes
- [ ] Ensure sufficient contrast
- [ ] Test in grayscale (for print)
- [ ] Keep legends clear and concise
- [ ] Remove unnecessary elements
- [ ] Consider your audience (presentation vs. report)

## Next Steps

- Review `vignette("color-palettes")` for palette selection
- See `vignette("plot-functions")` for helper functions
- Explore `?theme_benvi` for full documentation

## Further Reading

- [ggplot2 Theme Elements](https://ggplot2.tidyverse.org/reference/theme.html)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Data Visualization Best Practices](https://www.storytellingwithdata.com/)
